Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-03-22 _tex/paper2.tex
Built site for gh-pages

However, Thurstone designed Case V to provide a ``rather coarse
scaling'' of traits \citep[pp.~269]{Thurstone_1927b}, prioritizing
statistical simplicity over precise trait measurement. He cautioned that
its use ``should not be made without (an) experimental test''
\citep[pp.~270]{Thurstone_1927b} because it imposes the most extensive
set of simplifying assumptions among all the proposed cases
\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}}
(see Table~\ref{tbl-thurstone_cases}). Therefore, it is surprising that,
despite these limitations, CJ research has predominantly relied on Case
V to measure different traits, raising significant concerns about the
reliability and validity of such measurements in contexts where the
case's assumptions may not hold \citep[pp.~677]{Kelly_et_al_2022}.

@@

Despite its limitations, Case V remains the most widely used model in CJ
literature. This preference primarily stems from the BTL model, which
provides a simplified statistical representation of the case. The BTL
model mirrors the assumptions of Case V, with one notable distinction:
while Case V assumes a Normal distribution for the stimuli' discriminal
processes, the BTL model uses the more mathematically tractable Logistic
distribution \citep[pp.~254]{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). However, this substitution has minimal
impact on the model's estimation or interpretation because the
discriminal process scale is arbitrary up to a non-monotonic
transformation
\citetext{\citealp[pp.~16]{vanderLinden_et_al_2017_I}; \citealp{McElreath_2022}}.
Furthermore, this limited impact is supported by the fact that the
Normal and Logistic distributions exhibit analogous statistical
properties, differing only by a scaling factor of approximately \(1.7\).

@@

These discrepancies may arise from two possible causes. First, the
discriminal processes of misfit texts may substantially overlap with
those of non-misfit texts. Second, misfit texts may represent outlying
observations with characteristics that make them difficult to compare
and position on the trait scale. Regarding the first cause, the overlap
can occur when misfit texts share the same trait space and have equal
dispersion as non-misfit texts or when they exhibit larger discriminal
dispersions. Although the BTL model can distinguish the first scenario,
it is inherently agnostic to the second.
Figure~\ref{fig-discriminal_process} illustrates how an increased
discriminal dispersion can widen the overlap between two texts, leading
to more judgment discrepancies. In the figure, text B has a broader
discriminal dispersion than text A, expanding the overlap between their
discriminal processes, where judges struggle to determine which text
possesses more of the trait.

Thus, erroneously assuming equal dispersions between stimuli can give
rise to significant statistical and measurement issues. Specifically,
the BTL model may overestimate the trait's reliability, that is, the
degree to which the outcome accurately reflects the \emph{true}
discriminal differences between stimuli. This overestimation, in turn,
results in spurious conclusions about these differences
\citep[pp.~370]{McElreath_2020} and, by extension, about the underlying
discriminal processes of stimuli. Figure~\ref{fig-dispersion} also
illustrates this scenario when the model's discriminal difference
distribution aligns with the thick continuous line for
\(\sigma_{B}-\sigma_{A}=0\), while the \emph{true} discriminal
difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\).

@@

Finally, as discussed in the previous section, removing judges based on
misfit statistics risks discarding valuable information and even
introduce bias into the trait estimates
\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The direction and
magnitude of these biases remain unpredictable because they depend on
which judges researchers exclude from the analysis. Researchers define
\emph{misfit judges} as those whose evaluations deviate substantially
from the shared consensus due to the unique characteristics of either
the stimuli or the judges themselves
\citetext{\citealp[pp.~164-165]{Pollitt_2012a}; \citealp[pp.~289-290]{Pollitt_2012b}; \citealp[pp.~4]{vanDaal_et_al_2016}; \citealp[pp.~20]{Goossens_et_al_2018}}.

@@

Building on the previous section, it is clear that, despite its
limitations, researchers typically rely on the BTL model to measure a
trait and place the compared stimuli along its continuum
\citep[pp.~269]{Thurstone_1927b}. Additionally, the CJ literature shows
that researchers frequently use point estimates of BTL scores or their
transformations to conduct further analyses or hypothesis tests. For
example, researchers have used these scores to identify `misfit' judges
and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other assessment methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the underlying trait of interest
\citep{Casalicchio_et_al_2015, Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

Nevertheless, while separating the trait measurement and hypothesis
testing processes simplifies the analysis of CJ data, the statistical
literature cautions against relying solely on the point estimates of BTL
scores to conduct further analyses or hypothesis tests, as this practice
can undermine the resulting statistical conclusions. A key consideration
is that BTL scores are parameter estimates that inherently carry
uncertainty (measurement error). Ignoring this uncertainty can bias the
analysis and reduce the precision of hypothesis tests. The direction and
magnitude of such biases are often unpredictable. Results may be
attenuated, exaggerated, or remain unaffected depending on the degree of
uncertainty in the scores and the actual effects being tested
\citetext{\citealp[pp.~25]{Kline_et_al_2023}; \citealp[pp.~137]{Hoyle_et_al_2023}}.
Furthermore, the reduced precision in hypothesis tests diminishes their
statistical power, increasing the likelihood of committing type-I or
type-II errors \citep{McElreath_2020}.
Figure~\ref{fig-measurement_error} illustrates these issues, showing how
neglecting measurement error in an outcome can reduce the precision of a
predictor' estimated effect.

@@

A motivating example that will appear throughout the document clarifies
these concepts. In this example, researchers aim to determine: ``To what
extent do different teaching methods influence students' ability to
produce high-quality written texts?'' To investigate this, a researcher
designs a CJ experiment by randomly assigning students (individuals) to
two groups, each receiving a different teaching method. Judges then
compare pairs of students' written texts (stimuli) to produce a
dichotomous outcome reflecting the relative quality of each text
(trait). Based on this setup, researchers can reformulate the research
question as the estimand: ``\emph{On average}, is there a difference in
the ability to produce high-quality written texts between the two groups
of students?''. Following current CJ practices, researchers then rely on
estimates from the BTL model, or its transformations, to approximate
this estimand.

@@

Luckily, SCMs and DAGs support identification analysis through two key
advantages. First, regardless of complexity, they can represent various
causal structures using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
decompose complex structures into manageable components, facilitating
their analysis \citep{McElreath_2020}. Second, they depict causal
relationships in an interactive, non-parametric way. This flexibility
enables feasible identification strategies without requiring
specification of the types of variables, the functional forms relating
them, or the parameters of those functional forms
\citep[pp.~35]{Pearl_et_al_2016}.

@@

Accordingly, Section~\ref{sec-theory-theoretical_P} incorporates the
theoretical principles into what we refer to as the
\emph{conceptual-population model}. This model assumes that researchers
have access to a ``conceptual population'' of comparative data, that is,
data representing all repeated judgments made by every available judge
for each pair of stimuli produced by each pair of individuals in the
population. Conversely, Section~\ref{sec-theory-theoretical_SC}
integrates the assessment design features into what we refer to as the
\emph{sample-comparison model}. This model assumes a more realistic
scenario where researchers only have access to a sample of judges,
individuals, stimuli, and comparisons from the conceptual population.

@@

@@

Moreover, Figure~\ref{fig-cj03_dag} displays the DAG corresponding to
SCM~\ref{fig-cj03_scm}, illustrating the expected causal outlined in
Thurstone's theory. The graph shows that the discriminal processes of
the texts \((T_{ha}, T_{ib})\) influence their discriminal difference
\((D_{hiabjk})\), which in turn determines the outcome
\((O^{cp}_{hiabjk})\). It also highlights the influence of judges'
biases \((B_{kj})\) on the discriminal difference. Additionally, the DAG
differentiates between observed endogenous variables, such as the
outcome, and latent endogenous variables, including the texts'
discriminal processes, their discriminal difference, and the judges'
biases.

\subsubsection{The conceptual-population data
structure}\label{sec-theory-theoretical_P2}

Although specifying a system's data structure is not mandatory when
using SCMs and DAGs, defining one can improve clarity and facilitate the
description of the system. Thus, to re-express the scalar form of the CJ
system outlined in the previous section into an equivalent vectorized
form, we first define the vectors \(I\) and \(J\), along with the
matrices \(IA\) and \(JK\), as follows:

\begin{equation}\phantomsection\label{eq-mat12}{
I_{n_{I}} = \begin{bmatrix}
1 \\
2 \\
\vdots \\
h \\
\vdots \\
i \\
\vdots \\
n_{I}
\end{bmatrix} ; \;
J_{n_{J}} = \begin{bmatrix}
1 \\
2 \\
\vdots \\
j \\
\vdots \\
n_{J}
\end{bmatrix} ; \;
IA_{(n_{I} \cdot n_{A}) \times 2} = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{A}-1 \\
1 & n_{A} \\
\vdots & \vdots \\
h & a \\
\vdots & \vdots \\
i & b \\
\vdots & \vdots \\
n_{I} & n_{A}-1 \\
n_{I} & n_{A} \\
\end{bmatrix} ; \;
JK_{(n_{J} \cdot n_{K}) \times 2} = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{K}-1 \\
1 & n_{K} \\
\vdots & \vdots \\
j & k \\
\vdots & \vdots \\
n_{J} & n_{K}-1 \\
n_{J} & n_{K} \\
\end{bmatrix}
}\end{equation}

Here, each element of \(I\) represents a unique individual \(h\) or
\(i\), where \(n_{I}\) denotes the total number of individuals.
Similarly, each element of \(J\) corresponds to a unique judge \(j\),
with \(n_{J}\) indicating the total number of judges. Moreover, each row
of \(IA\) represents a unique pairing of individuals \(h, i\) with
stimuli \(a, b\). As a result, the matrix \(IA\) contains
\((n_{I} \cdot n_{A})\) rows and \(2\) columns, where \(n_{A}\)
specifies the number of stimuli available per individual. Likewise, each
row of \(JK\) associates a judge \(j\) with a judgment index \(k\),
where \(n_{K}\) indicates the number of judgments each judge makes.
Consequently, the matrix \(JK\) has \((n_{J} \cdot n_{K})\) rows and
\(2\) columns.

Additionally, we define the matrix \(RPUV\) that assigns each row from
the \(IA\) matrix to a corresponding row from the \(JK\) matrix. Thus,
the matrix consists of
\(n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\) rows and
\(6\) columns. Here, the term \({n_{I} \cdot n_{A} \choose 2}\)
represents the Binomial coefficient, which quantifies the total number
of repeated judgments each judge makes for every pair of stimuli
generated by each pair of individuals in the population. The matrix is
defined as follows:

\begin{equation}\phantomsection\label{eq-mat3}{
RPUV_{n \times 6} = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 1 & 2 & 1 & n_{K}-1 \\
1 & 1 & 1 & 2 & 1 & n_{K} \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
h & i & a & b & j & k \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 1 \\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K}-1 \\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K} \\
\end{bmatrix}
}\end{equation}

It is easier to visualize the structure of these vectors and matrices by
considering an example where \(n_{I} = 5\), \(n_{A} = 2\),
\(n_{J} = 3\), and \(n_{K} = 3\). In this simple case, the vectors and
matrices described in equations (\ref{eq-mat12}) and (\ref{eq-mat3})
take the following form:

\[
I_{5} = \begin{bmatrix}
1 \\
2 \\
3 \\
4 \\
5 \\
\end{bmatrix} ; \;
IA_{10 \times 2} = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 \\
\end{bmatrix} ; \;
J_{3} = \begin{bmatrix}
1 \\
2 \\
3 \\
\end{bmatrix} ; \;
JK_{9 \times 2} = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 \\
\end{bmatrix} ; \;
RPUV_{405  \times 6} = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 1 & 1 \\
1 & 1 & 5 & 1 & 1 & 2 \\
1 & 1 & 5 & 1 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 3 & 1 \\
1 & 1 & 5 & 1 & 3 & 2 \\
1 & 1 & 5 & 1 & 3 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
5 & 1 & 5 & 2 & 3 & 1 \\
5 & 1 & 5 & 2 & 3 & 2 \\
5 & 1 & 5 & 2 & 3 & 3 \\
\end{bmatrix}
\]

Then, using equations (\ref{eq-mat12}) and (\ref{eq-mat3}), we can
re-express SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in an
equivalent vectorized form, as shown in Figure~\ref{fig-cj04}. In this
depiction, the outcome \(O^{cp}_{RPUV}\), the texts' discriminal
difference \(D_{RPUV}\), their discriminal processes \(T_{IA}\), and the
judges' biases \(B_{JK}\) are represented as vectors rather than scalar
values. These vectors capture all the observations from the conceptual
population. Specifically, \(O^{cp}_{RPUV}\) is an observed vector of
length \(n={n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\), and
\(D_{RPUV}\) is a latent vector of the same dimension. Moreover,
\(T_{IA}\) and \(B_{JK}\) are latent vectors of lengths
\((n_{I} \cdot n_{A})\) and \((n_{J} \cdot n_{K})\), respectively.

@@

\[
\begin{aligned}
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK})
\end{aligned}
\]

@@

Building on the principles of Structural Equation Modeling (SEM)
\citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}},
the conceptual-population model incorporates two \emph{hierarchical
structural components}. A structural component describes how different
observed or latent variables influence the primary latent variable of
interest \citep{Everitt_et_al_2010}. Moreover, their hierarchical nature
enable analyses when the stimuli exhibit a hierarchical structure (refer
to Section~\ref{sec-theory-issue1b}) and when researchers aim to test
hypotheses while accounting for the uncertainties associated with the
traits (see Section~\ref{sec-theory-issue2}).

Figure~\ref{fig-cj09} displays these components applied to examine the
effect of different teaching methods on students' writing ability. The
top branch of DAG~\ref{fig-cj09_dag} depicts the first component. This
component describes how the students' writing-quality trait \((T_{I})\)
and a set of \emph{relevant} text-related variables \((X_{IA})\), such
as text length, causally influence the texts' written-quality trait
\((T_{IA})\). The DAG also highlights the idiosyncratic errors
\((e_{IA})\), which capture variations in the texts' trait that neither
the students' trait nor other relevant text-related variables can
explain. Here, \(T_{I}\) and \(e_{IA}\) are latent vectors of length
\(n_{I}\) and \((n_{I} \cdot n_{A})\), respectively. Additionally,
\(X_{IA}\) is an observed matrix with a dimension of
\((n_{I} \cdot n_{A})\) rows and \(q_{IA}\) columns.

Furthermore, the top branch of DAG~\ref{fig-cj09_dag} describes how a
set of \emph{relevant} student-related variables \((X_{I})\), including
teaching methods, causally influence the students' writing-quality trait
\((T_{I})\). The DAG also includes the idiosyncratic errors \((e_{I})\),
which capture variations in the students' traits that the relevant
student-related variables cannot explain. Here, \(e_{I}\) is a latent
vector of length \(n_{I}\), and \(X_{I}\) is an observed matrix with
\(n_{I}\) rows and \(q_{I}\) columns.

>>>>

However, Thurstone designed Case V to provide a ``rather coarse
scaling'' of traits \citep[pp.~269]{Thurstone_1927b}, prioritizing
statistical simplicity over precise trait measurement. He cautioned that
its use ``should not be made without (an) experimental test''
\citep[pp.~270]{Thurstone_1927b}, as it imposes the most extensive set
of simplifying assumptions among all the proposed cases
\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}}
(see Table~\ref{tbl-thurstone_cases}). Therefore, it is surprising that,
despite these limitations, CJ research has predominantly relied on Case
V to measure different traits, raising significant concerns about the
reliability and validity of such measurements in contexts where the
case's assumptions may not hold \citep[pp.~677]{Kelly_et_al_2022}.

@@

As previously discussed, Case V remains the most widely used model in CJ
literature. This preference primarily stems from the BTL model, which
provides a simplified statistical representation of the case. The BTL
model mirrors the assumptions of Case V, with one notable distinction:
while Case V assumes a Normal distribution for the stimuli' discriminal
processes, the BTL model uses the more mathematically tractable Logistic
distribution \citep[pp.~254]{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). However, this substitution has minimal
impact on the model's estimation or interpretation because the
discriminal process scale is arbitrary up to a non-monotonic
transformation
\citetext{\citealp[pp.~16]{vanderLinden_et_al_2017_I}; \citealp{McElreath_2021}}.
Furthermore, this limited impact is supported by the fact that the
Normal and Logistic distributions exhibit analogous statistical
properties, differing only by a scaling factor of approximately \(1.7\).

@@

These discrepancies may arise from two possible causes. First, the
discriminal processes of misfit texts may substantially overlap with
those of non-misfit texts. Second, misfit texts may represent outlying
observations with characteristics that make them difficult to compare
and position on the trait scale. Regarding the first cause, the overlap
can occur when misfit texts have equal dispersion and share the same
trait space as non-misfit texts or when they exhibit larger discriminal
dispersions. Although the BTL model can distinguish the first scenario,
it is inherently agnostic to the second.
Figure~\ref{fig-discriminal_process} illustrates how an increased
discriminal dispersion can widen the overlap between two texts, leading
to more judgment discrepancies. In the figure, text B has a broader
discriminal dispersion than text A, expanding the overlap between their
discriminal processes, a space where judges struggle to determine which
text possesses more of the trait.

When the assumption of equal dispersions between stimuli does not hold,
significant statistical and measurement issues can arise Specifically,
the BTL model may overestimate the trait's reliability, that is, the
degree to which the outcome accurately reflects the \emph{true}
discriminal differences between stimuli. This overestimation, in turn,
results in spurious conclusions about these differences
\citep[pp.~370]{McElreath_2020} and, by extension, about the underlying
discriminal processes of stimuli. Figure~\ref{fig-dispersion} also
illustrates this scenario when the model's discriminal difference
distribution aligns with the thick continuous line for
\(\sigma_{B}-\sigma_{A}=0\), while the \emph{true} discriminal
difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\).

@@

Finally, as discussed in the previous section, removing judges based on
misfit statistics risks discarding valuable information and even
introduce bias into the trait estimates
\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The direction and
magnitude of these biases remain unpredictable because they depend on
which judges researchers exclude from the analysis. \emph{Misfit judges}
are those whose evaluations deviate substantially from the shared
consensus due to the unique characteristics of either the stimuli or the
judges themselves
\citetext{\citealp[pp.~164-165]{Pollitt_2012a}; \citealp[pp.~289-290]{Pollitt_2012b}; \citealp[pp.~4]{vanDaal_et_al_2016}; \citealp[pp.~20]{Goossens_et_al_2018}}.

@@

Building on the previous section, it is clear that, researchers
typically rely on the BTL model to measure a trait and place the
compared stimuli along its continuum \citep[pp.~269]{Thurstone_1927b}.
Additionally, the CJ literature shows that researchers frequently use
point estimates of BTL scores or their transformations to conduct
further analyses or hypothesis tests. For example, researchers have used
these scores to identify `misfit' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other assessment methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the underlying trait of interest
\citep{Casalicchio_et_al_2015, Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

Nevertheless, while separating the trait measurement and hypothesis
testing processes simplifies the analysis of CJ data, the statistical
literature cautions against relying solely on the point estimates of BTL
scores to conduct further analyses or hypothesis tests, as this practice
can undermine the resulting statistical conclusions. A key consideration
is that BTL scores are parameter estimates that inherently carry
uncertainty (measurement error). Ignoring this uncertainty can bias the
analysis and reduce the precision of hypothesis tests. The direction and
magnitude of such biases are often unpredictable. Results may be
attenuated, exaggerated, or remain unaffected depending on the degree of
uncertainty in the scores and the actual effects being tested
\citetext{\citealp[pp.~25]{Kline_et_al_2023}; \citealp[pp.~137]{Hoyle_et_al_2023}}.
Furthermore, the reduced precision in hypothesis tests diminishes their
statistical power, increasing the likelihood of committing type-I or
type-II errors \citep{McElreath_2020}.
Figure~\ref{fig-measurement_error} illustrates these issues,
demonstrating how neglecting measurement error by relying only on
outcome averages can reduce the precision of a predictor's estimated
effect.

@@

A motivating example that will appear in the rest of the document
clarifies these concepts. In this example, researchers aim to determine:
``To what extent do different teaching methods influence students'
ability to produce high-quality written texts?'' To investigate this, a
researcher designs a CJ experiment by randomly assigning students
(individuals) to two groups, each receiving a different teaching method.
Judges then compare pairs of students' written texts (stimuli) to
produce a dichotomous outcome reflecting the relative quality of each
text (trait). Based on this setup, researchers can reformulate the
research question as the estimand: ``\emph{On average}, is there a
difference in the ability to produce high-quality written texts between
the two groups of students?''. Following current CJ practices,
researchers then rely on estimates from the BTL model, or its
transformations, to approximate this estimand.

@@

Luckily, SCMs and DAGs support identification analysis through two key
advantages. First, regardless of complexity, they can represent various
causal structures using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
decompose complex structures into manageable components, facilitating
their analysis \citep{McElreath_2020}. Second, they depict causal
relationships in a non-parametric way. This flexibility enables feasible
identification strategies without requiring specification of the types
of variables, the functional forms relating them, or the parameters of
those functional forms \citep[pp.~35]{Pearl_et_al_2016}.

@@

Accordingly, Section~\ref{sec-theory-theoretical_P} incorporates the
theoretical principles into what we refer to as the
\emph{conceptual-population model}. This model assumes that researchers
have access to a \emph{conceptual population} of comparative data, that
is, data representing all repeated judgments made by every available
judge for each pair of stimuli produced by each pair of individuals in
the population. Conversely, Section~\ref{sec-theory-theoretical_SC}
integrates the assessment design features into what we refer to as the
\emph{sample-comparison model}. This model assumes a more realistic
scenario where researchers only have access to a sample of judges,
individuals, stimuli, and comparisons from the conceptual population.

@@

In the conceptual-population model, we assume an idealized scenario
where researchers have access to a \emph{conceptual population} of
comparative data --data representing all repeated judgments made by each
judge for every stimulus pair produced by each pair of individuals in
the population. This assumption allows us to integrate Thurstone's
theoretical principles and propose innovations to address some of the
issues raised in Section~\ref{sec-theory-issues}.

@@

Figure~\ref{fig-cj03_dag} displays the DAG corresponding to
SCM~\ref{fig-cj03_scm}, illustrating the expected causal relationships
outlined in Thurstone's theory. The graph shows that the discriminal
processes of the texts \((T_{ha}, T_{ib})\) influence their discriminal
difference \((D_{hiabjk})\), which in turn determines the outcome
\((O^{cp}_{hiabjk})\). It also highlights the influence of judges'
biases \((B_{kj})\) on the discriminal difference. Additionally, the DAG
differentiates between observed endogenous variables, such as the
outcome (solid black circle), and latent endogenous variables, including
the texts' discriminal processes, their discriminal difference, and the
judges' biases (open circles).

\subsubsection{\texorpdfstring{The \emph{conceptual population} data
structure}{The conceptual population data structure}}\label{sec-theory-theoretical_P2}

Although specifying a system's data structure is not mandatory when
using SCMs and DAGs, defining one in this case can improve clarity and
facilitate the description of the system. Thus, to re-express the scalar
form of the CJ system shown in Figure~\ref{fig-cj03} into an equivalent
vectorized form, we first define the vectors \(I\) and \(J\), along with
the matrices \(IA\) and \(JK\), as follows:

\begin{equation}\phantomsection\label{eq-mat11}{
I = \begin{bmatrix}
1 \\
2 \\
\vdots \\
h \\
\vdots \\
i \\
\vdots \\
n_{I}
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
\vdots \\
j \\
\vdots \\
n_{J}
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{A}-1 \\
1 & n_{A} \\
\vdots & \vdots \\
h & a \\
\vdots & \vdots \\
i & b \\
\vdots & \vdots \\
n_{I} & n_{A}-1 \\
n_{I} & n_{A} \\
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{K}-1 \\
1 & n_{K} \\
\vdots & \vdots \\
j & k \\
\vdots & \vdots \\
n_{J} & n_{K}-1 \\
n_{J} & n_{K} \\
\end{bmatrix}
}\end{equation}

Here, each element of \(I\) represents a unique individual \(h\) or
\(i\), where \(n_{I}\) denotes the total number of individuals.
Similarly, each element of \(J\) corresponds to a unique judge \(j\),
with \(n_{J}\) indicating the total number of judges. Moreover, each row
of \(IA\) represents a unique pairing of individuals \(h, i\) with
stimuli \(a, b\). As a result, the matrix \(IA\) contains
\((n_{I} \cdot n_{A})\) rows and \(2\) columns, where \(n_{A}\)
specifies the number of stimuli available per individual. Likewise, each
row of \(JK\) associates a judge \(j\) with a judgment index \(k\).
Consequently, the matrix \(JK\) has \((n_{J} \cdot n_{K})\) rows and
\(2\) columns, where \(n_{K}\) indicates the number of judgments each
judge makes.

Additionally, we construct the matrix \(R\) to map each row of the
\(IA\) matrix with a corresponding row from the \(JK\) matrix. Thus,
this matrix has \(n\) rows and \(6\) columns, where
\(n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\). Here, the
term \({n_{I} \cdot n_{A} \choose 2}\) represents the binomial
coefficient, which quantifies the total number of unique comparisons
possible between every pair of stimuli generated by each pair of
individuals in the population. Whe define the matrix as follows:

\begin{equation}\phantomsection\label{eq-mat12}{
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 1 & 2 & 1 & n_{K}-1 \\
1 & 1 & 1 & 2 & 1 & n_{K} \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
h & i & a & b & j & k \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 1 \\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K}-1 \\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K} \\
\end{bmatrix}
}\end{equation}

It is easier to visualize the structure of these vectors and matrices by
considering an example where \(n_{I} = 5\), \(n_{A} = 2\),
\(n_{J} = 3\), and \(n_{K} = 3\). In this simple case, the vectors and
matrices described in equations (\ref{eq-mat11}) and (\ref{eq-mat12})
take the following form:

\[
I = \begin{bmatrix}
1 \\
2 \\
3 \\
4 \\
5 \\
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
3 \\
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 \\
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 \\
\end{bmatrix} ; \;
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 1 & 1 \\
1 & 1 & 5 & 1 & 1 & 2 \\
1 & 1 & 5 & 1 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 3 & 1 \\
1 & 1 & 5 & 1 & 3 & 2 \\
1 & 1 & 5 & 1 & 3 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
5 & 1 & 5 & 2 & 3 & 1 \\
5 & 1 & 5 & 2 & 3 & 2 \\
5 & 1 & 5 & 2 & 3 & 3 \\
\end{bmatrix}
\]

Now, using equations (\ref{eq-mat11}) and (\ref{eq-mat12}), we can
re-express SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in an
equivalent vectorized form, as shown in Figure~\ref{fig-cj04}. In this
depiction, the outcome \(O^{cp}_{R}\), the texts' discriminal difference
\(D_{R}\), their discriminal processes \(T_{IA}\), and the judges'
biases \(B_{JK}\) are represented as vectors rather than scalar values.
These vectors capture all the observations from the conceptual
population. Specifically, \(O^{cp}_{R}\) and \(D_{R}\) are observed and
latent vectors of length \(n\), respectively. Moreover, \(T_{IA}\) and
\(B_{JK}\) are latent vectors of lengths \((n_{I} \cdot n_{A})\) and
\((n_{J} \cdot n_{K})\), respectively.

@@

\[
\begin{aligned}
  O^{cp}_{R} & := f_{O}(D_{R}) \\
  D_{R} & := f_{D}(T_{IA}, B_{JK})
\end{aligned}
\]

@@

The conceptual-population model integrates two \emph{hierarchical
structural components} to examine how different teaching methods
influence students' writing ability. This approach builds on the
principles of Structural Equation Modeling (SEM)
\citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}}.
Each structural component defines how observed or latent variables
affect the primary latent variable of interest
\citep{Everitt_et_al_2010}. Their hierarchical nature enables
researchers to test hypotheses while accounting for uncertainties
associated with trait estimation (see Section~\ref{sec-theory-issue2}),
particularly when stimuli exhibit a hierarchical structure (see
Section~\ref{sec-theory-issue1b}).

The top branch of DAG~\ref{fig-cj09_dag} depicts the first component.
This component describes how the students' writing-quality trait
\((T_{I})\) and a set of \emph{relevant} text-related variables
\((X_{IA})\) (e.g., text length) causally influence the texts'
written-quality trait \((T_{IA})\). The DAG also highlights the
idiosyncratic errors \((e_{IA})\), which capture variations in the
texts' trait that neither the students' trait nor other relevant
text-related variables can explain. In this structure, \(T_{I}\) and
\(e_{IA}\) are latent vectors of length \(n_{I}\) and
\((n_{I} \cdot n_{A})\), respectively. Additionally, \(X_{IA}\) is an
observed matrix with a dimension of \((n_{I} \cdot n_{A})\) rows and
\(q_{IA}\) columns (variables).

Additionally, the DAG's top branch illustrates how a set of
\emph{relevant} student-related variables \((X_{I})\), such as the
teaching method, causally influence the students' writing-quality trait
\((T_{I})\). The DAG also includes the idiosyncratic errors \((e_{I})\),
which capture variations in the students' traits that the relevant
student-related variables cannot explain. Here, \(e_{I}\) is a latent
vector of length \(n_{I}\), and \(X_{I}\) can be an observed matrix with
\(n_{I}\) rows and \(q_{I}\) columns (variables).
